---
title: "Code for reproducibility of presentation plots"
subtitle: "Exam Number: B277001"
output: 
  prettydoc::html_pretty:
    theme: leonids
    toc: yes
date: "2024-11-30"
---

```{r setup, include=TRUE, message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, tidy.opts = list(width.cutoff = 60), tidy = TRUE)

if (!require("pacman")) install.packages("pacman")
pacman::p_load(bookdown, dplyr, tidyr, ggplot2, ggpubr, ggrepel, ggthemes, gt, janitor, MetBrewer,pander, patchwork, plotly, prettydoc,readr, readxl, sf, skimr, slider, stringr, tidyselect)
```


## International data, source: OECD 

### Data load and preparation

```{r oecd-screening-load}
# data source: https://data-explorer.oecd.org/vis?fs[0]=Topic%2C1%7CHealth%23HEA%23%7CHealthcare%20use%23HEA_PRC%23&pg=0&fc=Topic&bp=true&snb=10&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_HEALTH_PROC%40DF_SCREEN&df[ag]=OECD.ELS.HD&df[vs]=1.1&dq=.................&pd=2010%2C&to[TIME_PERIOD]=false
overall_screening <- read.csv("data/0_screening_oecd_24112024.csv")

overall_screening %>% 
  filter(DATA_SRCE == "PRG",
         #using <2015 to match the availability of OECD cancer survival data 
         TIME_PERIOD < 2015) %>% 
  mutate(year = as.numeric(TIME_PERIOD), 
         country = factor(Reference.area),
         value = OBS_VALUE) %>% 
  group_by(Cancer.site, country) %>% 
  #keep countries that have  at least 12 records for the period 2000-2014 for each cancer site
  filter(length(country) >= 12) %>% 
  ungroup() %>%
  #calculate slide window with 5 years lag
  mutate(value_5y = slider::slide_dbl(value, mean, .before = 4, .after = 0),
         #create a name_lab variable for plot labels
         name_lab = if_else(year == 2014, country, NA_character_)) %>% 
  #filter by years matching the cancer survival records
  filter(year %in% c(2004, 2009, 2014)) -> df_plt_screening

#generate a named color vector for all countries
colour_vec_all <- met.brewer("VanGogh2", length(unique(df_plt_screening$country)))
names(colour_vec_all)<- unique(df_plt_screening$Reference.area)

```

### Data subset for breast neoplasms 

```{r subsetting-breast}
cancer_site = 'Malignant neoplasms of female breast'
colour_vec = colour_vec_all[as.character(unique(pull(df_plt_screening[df_plt_screening$Cancer.site == cancer_site, ],'country')))] 

#
screening_breast <- df_plt_screening %>% 
  filter(Cancer.site == cancer_site) %>% 
  select(year, country, value_5y) %>%
  # create a composed year_country to match with mortality records
  mutate(year_country = interaction(year, country, sep = '-')) 

```

### Mortality data for breast neoplasms 

```{r oecd-mortality-load}
# data source:https://data-explorer.oecd.org/vis?fs[0]=Topic%2C1%7CHealth%23HEA%23%7CHealthcare%20quality%20and%20outcomes%23HEA_HCQ%23&pg=0&fc=Topic&bp=true&snb=11&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_HCQO%40DF_CC&df[ag]=OECD.ELS.HD&df[vs]=1.0&dq=AUT%2BAUS.A.CCCLNTSR..Y_GE15._T.OBS&pd=2004%2C&to[TIME_PERIOD]=false
overall_inc_mor <- read.csv("data/1_survival_oecd_24112024.csv")

mortality_breast <- overall_inc_mor %>% 
  filter(Measure == 'Breast cancer five-year net survival',
         Statistical.operation == 'Observed') %>% 
  mutate(year = TIME_PERIOD, 
         country = Reference.area,
         year_country = interaction(year, country, sep = '-'),
         mortality = OBS_VALUE) %>% 
  select(year, country, mortality, year_country)
```

### Join screening and mortality datasets 

```{r join-breast-data}
breast_join <- left_join(screening_breast, 
                         mortality_breast, 
                         by = 'year_country')

#get color vector for the countries represented
colour_vec <- colour_vec_all[as.character(unique(pull(breast_join,'country.x')))] 
```


### Density plot Slide 2, panel B 

```{r slide2-plotB}
screening_breast %>% 
  mutate(year = case_when(year == 2004 ~"1999-2004",
                          year == 2009 ~ "2005-2009",
                          year == 2014 ~ "2010-2014"),
         year = factor(year)) %>% 
  ggplot(aes(x = value_5y, color = year, fill = year)) +
  geom_density(alpha = 0.6) +
  geom_segment(data = screening_breast %>% 
                 mutate(year = case_when(year == 2004 ~"1999-2004",
                                         year == 2009 ~ "2005-2009",
                                         year == 2014 ~ "2010-2014"),
                        year = factor(year)) %>% 
                 group_by(year) %>% 
                 summarise(median_screening = median(value_5y)) %>% 
                 ungroup(), 
               aes(x = median_screening,
                   xend = median_screening,
                   y = 0,
                   yend = Inf),
               colour= 'grey30')+
  geom_text(data = screening_breast %>% 
              mutate(year = case_when(year == 2004 ~"1999-2004",
                                      year == 2009 ~ "2005-2009",
                                      year == 2014 ~ "2010-2014"),
                     year = factor(year)) %>% 
              group_by(year) %>% 
              summarise(median_screening = median(value_5y)) %>% 
              ungroup(),
            aes(x = 30, y = 0.03,label = paste0('median = ',median_screening, "%")),
            fontface = 'bold',
            size = 10, size.unit = "pt") +
  scale_fill_viridis_d(direction = -1, end = 0.9) +
  scale_color_viridis_d(direction = -1, end = 0.9) +
  theme(legend.position="none") +
  labs(title = 'Screening of malignant neoplasms of the breast',
       subtitle = "Target population: Female, 50-69 years old. Source: OECD (https://data-explorer.oecd.org/).",
       x = 'Screening (% target population)',
       y = 'Density') +
  guides(fill = 'none', colour= 'none') +
  facet_wrap(.~year) +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, hjust = 0.5,face = "bold"),
        plot.title = element_text(size = 15, hjust = 0.5,face = "bold"),
        plot.subtitle = element_text(size=8,hjust = 0.5,face = "bold")) -> density_breast

print(density_breast)

#uncomment lines below for saving

# png("Rplots/density_breast.png",res = 300,units = 'in',height = 3, width = 6)
# print(density_breast)
# dev.off()
```


### Scatter and line plot Slide 2, panel C 

```{r slide2-plotC}
breast_join %>% 
  mutate(year = as.numeric(year.x)) %>%
  rename(Country = country.x,
         Screening = value_5y,
         Survival = mortality, ) %>% 
  mutate(name_lab = if_else(year == 2009, Country, NA_character_)) %>% 
  ggplot(aes(x= Screening, y = Survival)) +
  geom_path(aes(group = Country, colour = Country),
            alpha = 1,
            linewidth = 1,
            linejoin = 'round')+
  geom_text_repel(aes(color = Country, label = name_lab),
                  fontface = "bold",
                  size = 5,
                  direction = "y",
                  xlim = c(95, NA),
                  hjust = 0,
                  segment.size = .7,
                  segment.alpha = 0,
                  box.padding = .4,
                  segment.curvature = -0.1,
                  segment.ncp = 3,
                  segment.angle = 20)+
  geom_point(size = 0.5, shape = 16) +
  geom_text_repel(aes(label = year),
    size = 3,
    direction = "both",
    force_pull = 10,
    segment.size = 0.5,
    segment.alpha = 1,
    segment.linetype = "dotted",
    box.padding = 0.2,
    segment.curvature = -0.11,
    segment.ncp = 3,
    segment.angle = 20) +
  scale_colour_manual(values = colour_vec) +
  scale_x_continuous(limits = c(30,100),
                     breaks = seq(30,100, 10),
                     expand = expansion(mult=c(0.05,.12)))+
  scale_y_continuous(limits = c(65,95),
                     breaks = seq(65,95, 15))+
  theme_minimal() +
  labs(title = 'Screening and patient\'s survival over time for breast neoplasms',
       x = 'Screening (% target population)',
       y='Net Survival (% of population with disease)') +
  theme(plot.title = element_text(size = 15, hjust = 0.5,face = "bold"),
        plot.subtitle = element_text(size=8,hjust = 0.5,face = "bold"),
        axis.title =  element_text(size = 12, hjust = 0.5,face = "bold")) +
  guides(colour='none')-> breast_mortalityscreening

print(breast_mortalityscreening)


#uncomment lines below for saving

# png(filename = 'Rplots/breast_plt.png',res = 300,width = 6.8, height = 5.8, units = 'in')
# print(breast_mortalityscreening)
# dev.off()
```
## Scotland data, source: Public Health Scotland

```{r load-scot-phs}
#read file dictionary of health boards names and codes
# source: https://geoportal.statistics.gov.uk/documents/844159d820da487686d124a25e2eb84d/about
hb21 <- read.csv("data/2_hbscot_codes_name.csv")

#read a simple geojson to generate the map 
# source: https://www.arcgis.com/apps/mapviewer/index.html?layers=52aecbed981b42e6ab4030510873e6dd
map_HB_scot <- st_read("data/3_hbscot_map.geojson")

# read mortality data Scotland per Health Board
# source: https://www.opendata.nhs.scot/dataset/cancer-mortality/resource/f0446372-e55a-441e-ae71-0b21a1cdda2c
mort <- read.csv("data/4_mortality_hbScot_24112024.csv")


# filter mortality for breast cases
mort_filter <- mort %>%
  filter(Year == '2016-2020', 
         CancerSite == 'Breast', 
         Sex == 'Female')

#merge mortality dataset with dictionary
mort_filter_hb21 <- merge(hb21, mort_filter, by.x = 'HB21CD', by.y = 'HB') %>%
  # adding \n so long HB names with & get split
  mutate(HB21NM = str_replace_all(HB21NM," and ", "\n & " ))

map_HB_scot. <- map_HB_scot %>% 
  mutate(h06_name = str_replace_all(h06_name," & ", "\n & " ))

# merge mortality and the dataframe from the map by Health Board name
map_HBmortality_scot <- merge(map_HB_scot, mort_filter_hb21, by.x = 'h06_name', by.y = 'HB21NM')
```
### Map plot slide 4, panel A

```{r plot-map}
# Plot with ggplot2's sf geom
ggplot(data = map_HBmortality_scot, mapping = aes(label=h06_name)) +
  #map the world-age standardised rate (WASR) of deaths
  geom_sf(aes(fill = WASR), color='white') +
  scale_fill_fermenter(direction = 1, palette = "GnBu")+  
  geom_sf_text(fontface = "bold",size = 2)+
  labs(title = "Breast cancer mortality in Scotland",
       subtitle = "World age-standardised mortality rate (WASR) per 100,000 person-years at risk \n for period 2016-2020 across all cancer types by Health Board.",
       caption = "Source: Public Health Scotland (https://www.opendata.nhs.scot).") +  
  theme_map()+
  theme(
    legend.position = "top",
    legend.justification = 0.5,
    legend.key.size = unit(0.75, "cm"),
    legend.key.width = unit(1.0, "cm"),
    legend.text = element_text(size = 8),
    legend.margin = margin(),
    plot.title = element_text(size = 15, hjust = 0.5,face = "bold"),
    plot.subtitle = element_text(size=8,hjust = 0.5,face = "bold")
  )-> breast_map

print(breast_map)

#uncomment lines below for saving
# png("Rplots/map_breast_scot.png",res = 300,units = 'in',height = 6, width = 5)
# print(breast_map)
# dev.off()
```
### Breast screening in Scotland. Figure slide 4, panel B

```{r scot-breast-screening, message=F}
# data load and wrangling
#data source: 
screening_scotland <- read_excel("data/5_screening_hbscot_2223.xlsx")
hb_names <- str_replace_all(map_HB_scot.$h06_name, '\n', '')

colour_vec_hb <- met.brewer("VanGogh2", length(hb_names))
names(colour_vec_hb)<- hb_names

#add color black for national level screening rates
colour_vec_hb_scot = c(colour_vec_hb,  setNames("black", "Scotland"))
#define line type vector, segmented for national level
lty_hb_scot <- c(rep(1, length(hb_names)), 2)
names(lty_hb_scot)<- names(colour_vec_hb_scot)
```

```{r plot-scotl-breast-screening}
screening_scotland %>% 
  select(`Scottish Breast Screening Programme`:...12) %>% 
  filter(`Scottish Breast Screening Programme` %in% c(hb_names, "Scotland","NHS Board of Residence")) %>% 
  row_to_names(row_number = 1) %>% 
  select(where( ~!all(is.na(.x)))) %>%
  pivot_longer(names_to = "Period", 
               cols = `2011/14`:`2020/23`,
               values_to = 'Uptake') %>% 
  rename(hb_name = `NHS Board of Residence`) %>% 
  separate_wider_delim(Period,"/", names = c(NA, "Period")) %>% 
mutate(Uptake = as.numeric(Uptake),
       Period = as.numeric(Period),
       hb_name = factor(hb_name),
       hb_name_label = ifelse(Period == 23, as.character(hb_name), NA_character_),
       #hb_name_label = str_replace_all(hb_name_label, "&", "\n&")
       ) %>% #View()
  ggplot(aes(Period, Uptake, group = hb_name, 
             linetype = hb_name,
             label = hb_name_label)) +
  geom_line(aes(colour = hb_name))+
  geom_text_repel(aes(label = str_replace_all(hb_name_label, "&", "\n&"), 
                      colour = hb_name_label),
                  fontface = "bold",
                  size = 3.8,
                  direction = "y",
                  segment.alpha = 0,
                  xlim = c(23.5, NA),
                  hjust = 0,
                  box.padding = .45) +
  scale_colour_manual(values = colour_vec_hb_scot)+
  scale_linetype_manual(values = lty_hb_scot) +
  #show target screenign rate at 80%
  geom_segment(aes(x = 14, xend = 23, y = 80, yend = 80), colour= 'black') +
  geom_label(aes(x = 19, y=80, label="Target"), fill='white') +
  scale_x_continuous(expand=expand_scale(mult=c(0.05,.6)),
                     labels = paste0("20",seq(11,20),"-20",seq(14,23)),
                     breaks = seq(14,23)) +
  guides(colour = 'none', linetype = 'none')+
  labs(title = 'Screening of malignant neoplasms of the breast',
       subtitle = "Health Board screening over three-year rolling period. \nTarget population: Female, 50-70 years old.",
       caption = "Source: Public Health Scotland (https://publichealthscotland.scot/)",
       y = 'Screening (% target population)',
       x = 'Three-year rolling period') +
  theme_minimal() +
  theme(plot.title = element_text(size = 15, hjust = 0.5,face = "bold"),
        plot.subtitle = element_text(size = 8,hjust = 0.5,face = "bold"),
        axis.text.x = element_text(angle = 45)) -> scot_screen_breast

print(scot_screen_breast)

#uncomment next lines for saving

# png("Rplots/screening_breast_scot.png",res = 300,units = 'in',height = 8, width = 8)
# print(scot_screen_breast)
# dev.off()
```
### Annex: Session info
```{r session-info, echo=FALSE}
pander(sessionInfo())
```